---
title: "Untitled"

---
Analysis of the Washignton State crash data
```{r}
getwd()
```
```{r}
#setwd()
```
cit <- read.csv("citations.CSV", quote = "", 
                 row.names = NULL, 
                 stringsAsFactors = FALSE)
```{r}
raw_data <- read.csv(
    "WA_Crash_Summary.csv",
    header = TRUE,
    sep = ",",
    quote = "",
    row.names = NULL,
    stringsAsFactors = FALSE
    )
```
```{r}
dim(raw_data)
```

```{r}
head(raw_data)
```
Want the datatype of a column
```{r}
typeof(raw_data$Damage.Threshold.Met)
```
Column names of the data frame.
```{r}
colnames(raw_data)
```
Get unique categories in a column
```{r}
#unique(raw_data[, 15])
unique(raw_data$X.Collision.Type.)
```
Want to see the different cities in the appropriate column
```{r}
unique(raw_data$X.Injury.Severity.)
```
Import necessary library
```{r}
library(dplyr) # for functions
```
```{r}
raw_data %>% filter(X.Injury.Severity. ==  "\"Daylight\"")
```
Will import data.table package to make importing of CSV file faster
```{r}
library(data.table)
```
```{r}
df <- fread("/Volumes/my_passport/data_sets/WA_Crash_Summary.csv")
```
```{r}
categories <- unique(yourDataFrame$yourColumn) 
numberOfCategories <- length(categories)
```