---
title: "Exploratory Data Analysis of Washington State Crash Data"
author: J. Bravo
date: "December 4, 2023"
---

Import necessary libraries.
Will import data.table package to make importing of CSV file faster.
```{r}
library(dplyr) # for functions
library(data.table)
```
Reside in same directory as the dataset(s).
```{r}
getwd()
```
Import the dataset.
```{r}
raw_data <- fread("/Volumes/my_passport/data_sets/WA_Crash_Summary.csv")
```
Size of the dataframe.
```{r}
dim(raw_data)
```
According to the command above, this sample is 330k rows by 25 columns.
Want a sample of the first six entries.
```{r}
head(raw_data)
```
Datatype check.
```{r}
print(class(raw_data))
```
Dataype check on a column.
```{r}
print(class(raw_data$"Pedestrians Involved"))
print(class(raw_data$"Pedalcyclists Involved"))
```
Column names of the data frame.
```{r}
colnames(raw_data)
```
Get unique categories in a column
```{r}
unique(raw_data$"City")
unique(raw_data$"Hit and Run")
```
The cities column has 281 unique entries, whereas the Hit and Run column is boolean T or F.
Store the number of records in its own variable for later analysis.
```{r}
num_of_records <- nrow(raw_data)
num_of_records
```
Check the unique values in Injury Severity.
```{r}
unique(raw_data$"Injury Severity")
```
Want new dataframe to hold those records where only "Unknown Injury Collision" is in the "Injurity Severity" field.
```{r}
unknowns <- raw_data %>%
    filter(raw_data$"Injury Severity" == "Unknown Injury Collision")
num_of_unknowns <- nrow(unknowns)
```
Get proportion of entries that have "Unknown Injury Collision" status
```{r}
num_of_unknowns / num_of_records
```
So according to above results, about 10% of collisions are categorized as Unknown.
Similar analysis for the proportion of accidents categorized as "Serious Injury Collision"
```{r}
serious <- raw_data %>%
    filter(raw_data$"Injury Severity" == "Serious Injury Collision")
num_of_serious <- nrow(serious)
num_of_serious / num_of_records
```
So only about a percent of the collisions have been reported as Serious.
Following is used to get the number of unique caegories within a column.
```{r}
categories <- unique(yourDataFrame$yourColumn)
numberOfCategories <- length(categories)
```