---
title: "Exploratory Data Analysis"
author: "J. Bravo"
date: "2024-01-06"
output: html_document
---
Initial data analysis to investigate a new crash data set.
Load the necessary libraries.
```{r}
library(dplyr) # for functions
library(data.table)
library(ggplot2)
```
Load the data.It is Montgomery County Crash Reporting - Incidents Data
```{r}
raw_data <- fread("C:/Users/joepb/Desktop/unzipfolder/mont_cty_crash_reporting_incidents_data.csv")
```
Want to see the first five values in this initial dataframe.
```{r}
head(raw_data)
```
Want to get the number of fields in this data frame. One has to serve as the predictor (dependent) variable and the other will serve as predictor (independent) variables. But not all columns will be used.
```{r}
ncol(raw_data)
```
#### ACRS Report Type
Get the unique values in the "ACRS Report Type" column. This may be the outcome (dependent) variable.
```{r}
unique(raw_data$"ACRS Report Type")
```
This is a consideration for the outcome variable. Will consider some more before investigating more.
#### Collision Type
```{r}
unique(raw_data$"Collision Type")
```
This is a consideration for the predictor (independent) variable.
#### Collision Type
```{r}
unique(raw_data$"Injury Severity")
```
This field has five possible outcomes as compared to the three in the ACRS Report Type. This may be a better option for outcome variable.
#### Vehicle Damage Extent
```{r}
unique(raw_data$"Vehicle Damage Extent")
```
#### Possible Dependent Variable - Vehicle Damage Extent
```{r}
unique(raw_data$"Vehicle Damage Extent")
```
#### Possible Predictor Variable - Vehicle Body Type
```{r}
unique(raw_data$"Vehicle Body Type")
```
#### Possible Predictor Variable - Vehicle Movement
```{r}
unique(raw_data$"Vehicle Movement")
```
#### Possible Outcome Variable - ACRS Report Type
```{r}
unique(raw_data$"ACRS Report Type")
```
#### Possible Predictor Variable - Speed Limit
```{r}
unique(raw_data$"Speed Limit")
```
#### Possible Predictor Variable - Latitude
```{r}
tail(raw_data$"Latitude")
```

## Weather
Want to look at the exploratory analysis for this qualitative variable.
Look at the unique values in this field.
```{r}
unique(raw_data$"Weather")
```
### Frequency Table
Now, want frequency distribution of qualitative data
```{r}
freq_tbl_weather <- table(raw_data$"Weather")
print("Frequency Table")
print(freq_tbl_weather)
```
### Plots
Plot frequency distribution using ggplot2. Note! The following dataframe holds the frequency counts. It does not hold any more information from the original dataframe.
```{r}
# Convert from table to df
freq_df_weather <- as.data.frame(freq_tbl_weather)
# Plotting logic
ggplot(freq_df_weather, aes(x = Var1, y = Freq)) +
  geom_point(colour = "firebrick3") +
  labs(title = "Weather Conditions Distribution", x = "Weather", y = "Frequency") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
A code snippet on how to convert character to factor data types.
```{r}
freq_tbl_weather <- raw_data %>%
  mutate_if(is.character, as.factor)
freq_tbl_weather
```
Plot frequency distribution using ggplot2.
```{r}
# Plotting logic
ggplot(raw_data, aes(x = as.factor(raw_data$"Weather"))) +
  geom_bar(fill = "cadetblue2", color = "white") +
  labs(title = "Weather Conditions Distribution", x = "Weather", y = "Counts") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
### Filtering
Before performing Chi-Square testing for the predictor variables, must handle empty strings and N/A's.
Will first remove the N/A's from this dataset.
```{r}
weather_df <- raw_data %>%
    filter(raw_data$"Weather" == "N/A")
nrow(weather_df)
```
The number of rows matches the results from the earlier frequency table.
### Chi-Square Testing
This next chi-square test will check the relationship between injury severity and weather.
```{r}
weather_tbl <- table(weather_df$"Injury Severity", weather_df$"Weather")
print(weather_tbl)
print(chisq.test(weather_tbl, simulate.p.value = TRUE))
```
The resulting p-value = 0.0004998 is less than the 0.05 threshold. Hence, I reject the
Null Hypothesis that weather and injury severity are independent of each other.
Will need to implement some oversampling or undersampling to balance the dataframe. The Clear class within the Weather field is over-represented.

## Light
Another qualitative variable
Unique values in this column.
```{r}
unique(raw_data$"Light")
```
### Frequency Table
Now, want frequency distribution of qualitative data
```{r}
freq_tbl_light <- table(raw_data$"Light")
print("Frequency Table")
print(freq_tbl_light)
```
### Plots
Plot frequency distribution.
```{r}
# Convert from table to df
freq_df_light <- as.data.frame(freq_tbl_light)
# Plotting logic
ggplot(freq_df_light, aes(x = Var1, y = Freq)) +
  geom_point(colour = "firebrick3") +
  labs(title = "Lighting Conditions Distribution", x = "Lighting", y = "Frequency") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
Plot frequency distribution using ggplot2.
```{r}
# Plotting logic
ggplot(raw_data, aes(x = as.factor(raw_data$"Light"))) +
  geom_bar(fill = "cadetblue2", color = "white") +
  labs(title = "Lighting Conditions Distribution", x = "Lighting", y = "Counts") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
### Filtering
Before performing Chi-Square testing for the predictor variables, must handle empty strings and N/A's.
Will first remove the N/A's from this dataset.
```{r}
light_df <- raw_data %>%
    filter(raw_data$"Light" != "N/A",
           raw_data$"Light" != "OTHER",
           raw_data$"Light" != "UNKNOWN")
nrow(light_df)
```
The number of rows matches the results from the earlier frequency table.
### Chi-Square Testing
This next chi-square test will check the relationship between injury severity and lighting.
```{r}
light_tbl <- table(light_df$"Injury Severity", light_df$"Light")
print(light_tbl)
print(chisq.test(light_tbl, simulate.p.value = TRUE))
```
The resulting p-value = 0.0004998 is less than the 0.05 threshold. Hence, I reject the
Null Hypothesis that light and injury severity are independent of each other.
Will need to implement some oversampling or undersampling to balance the dataframe. The Daylight class within the Light field is over-represented.

## Vehicle Body Type
Third and final qualitative variable.
Unique values within the field.
```{r}
unique(raw_data$"Vehicle Body Type")
```
### Frequency Table
Frequency distribution of Vehicle Body Type.
```{r}
# vbt <-- vehicle body type
freq_tbl_vbt <- table(raw_data$"Vehicle Body Type")
print("Frequency Table")
print(freq_tbl_vbt)
```
### Plots
Frequency distribution plots using ggplot2.
```{r}
# Convert from table to df
freq_df_vbt <- as.data.frame(freq_tbl_vbt)

# Logic
labels = c(suv = "(SPORT) UTILITY VEHICLE",
           blank = "",
           atv = "ALL TERRAIN VEHICLE (ATV)",
           amb_em = "AMBULANCE/EMERGENCY",
           amb_nem = "AMBULANCE/NON EMERGENCY",
           au_cyl = "AUTOCYCLE",
           two_axl = "CARGO VAN/LIGHT TRUCK 2 AXLES (OVER 10,000LBS (4,536 KG))",
           ccb = "CROSS COUNTRY BUS",
           fv = "FARM VEHICLE",
           fv_em = "FIRE VEHICLE/EMERGENCY",
           fv_nem = "FIRE VEHICLE/NON EMERGENCY",
           limo = "LIMOUSINE",
           lsv = "LOW SPEED VEHICLE",
           three_axl = "MEDIUM/HEAVY TRUCKS 3 AXLES (OVER 10,000LBS (4,536KG))",
           mpd = "MOPED",
           mc = "MOTORCYCLE",
           na = "N/A",
           oth = "OTHER",
           oth_bs = "OTHER BUS",
           olt = "OTHER LIGHT TRUCKS (10,000LBS (4,536KG) OR LESS)",
           pass = "PASSENGER CAR",
           pickup = "PICKUP TRUCK",
           pol_em = "POLICE VEHICLE/EMERGENCY",
           pol_nem = "POLICE VEHICLE/NON EMERGENCY",
           rv = "RECREATIONAL VEHICLE",
           sb  = "SCHOOL BUS", 
           sn_mob = "SNOWMOBILE",
           sw = "STATION WAGON",
           tr_bus = "TRANSIT BUS",
           tt = "TRUCK TRACTOR",
           unk = "UNKNOWN",
           vn = "VAN"
           )

# Plotting logic
ggplot(data = freq_df_vbt, aes(x=Var1, y=Freq)) +
  geom_point(colour = "firebrick3") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  geom_text(aes(label = Var1), hjust = -0.05, angle = 45, size = 2.25)
```
### Filtering
Before performing Chi-Square testing for the predictor variables, must handle empty strings, N/A's, and "others".
Will first remove the N/A's from this dataset.
```{r}
vbt_df <- raw_data %>%
    filter(raw_data$"Vehicle Body Type" != "N/A",
           raw_data$"Vehicle Body Type" != "OTHER",
           raw_data$"Vehicle Body Type" != "UNKNOWN",
           raw_data$"Vehicle Body Type" != ""
           )
nrow(vbt_df)
```
The number of rows matches the results from the earlier frequency table.
### Chi-Square Testing
This next chi-square test will check the relationship between injury severity and lighting.
```{r}
vbt_tbl <- table(vbt_df$"Injury Severity", vbt_df$"Vehicle Body Type")
print(vbt_tbl)
print(chisq.test(vbt_tbl, simulate.p.value = TRUE))
```
The resulting p-value = 0.0004998 is less than the 0.05 threshold. Hence, I reject the
Null Hypothesis that vehicle body type and injury severity are independent of each other.
Will need to implement some oversampling or undersampling to balance the dataframe. The passenger car class within the vehicle body type field is over-represented.

 




























