---
title: "Smote implementation in unbalanced data"
author: "J. Bravo"
date: "2023-12-20"
output: html_document
---
Load the necessary libraries.
```{r}
library(dplyr) # for functions
library(data.table)
library(janitor)
```
Load the dataset.
```{r}
#raw_data <- fread("C:/Users/joepb/Desktop/unzipfolder/WA_Crash_Summary.csv")
raw_data <- fread("/Volumes/my_passport/data_sets/WA_Crash_Summary.csv")
```
Check a sample
```{r}
head(raw_data)
```
Install the necessary library and include the library.
```{r}
install.packages("SMOTE")
#library(SMOTE)
```

Get unique values from data frame.
```{r}
unique(raw_data$"Lighting Condition")
```
```{r}
unique(raw_data$"Weather Condition")
```

Get count of empty string records in the Lighting Condition and Weather Condition fields.
```{r}
raw_data %>%
  filter(raw_data$"Lighting Condition" == "") %>%
  count()
```
```{r}
raw_data %>%
  filter(raw_data$"Weather Condition" == "") %>%
  count()
```

Now that I've seen how to remove the empty strings in each column, I will implement in one chunk.
```{r}
initial_df <- raw_data %>%
  filter(
    raw_data$"Weather Condition" != "",
    raw_data$"Lighting Condition" != ""
    ) %>%
  select(
    "City",
    "Weather Condition",
    "Lighting Condition",
    "Injury Severity"
    )
# Change the column names
colnames(initial_df) <- c(
    "city",
    "weather_condition",
    "lighting_condition",
    "injury_severity"
    )
nrow(initial_df)
```

Store the number of records in its own variables.
```{r}
num_records <- nrow(initial_df)
num_records
```
Get proportions of the other classes
```{r}
prop1 <- nrow(
  initial_df %>%
  filter(
    initial_df$injury_severity == "No Injury Collision")
    ) / num_records
prop2 <- nrow(
  initial_df %>%
  filter(
    initial_df$injury_severity == "Minor Injury Collision")
    ) / num_records
prop3 <- nrow(
  initial_df %>%
  filter(
    initial_df$injury_severity == "Unknown Injury Collision")
    ) / num_records
prop4 <- nrow(
    initial_df %>%
    filter(
        initial_df$injury_severity == "Serious Injury Collision")
    ) / num_records
prop5 <- nrow(
    initial_df %>%
    filter(
        initial_df$injury_severity == "Fatal Collision")
    ) / num_records
print(paste("Proportion for No Injury Collision", prop1))
print(paste("Proportion for Minor Injury Collision", prop2))
print(paste("Proportion for Unknown Injury Collision", prop3))
print(paste("Proportion for Serious Injury Collision", prop4))
print(paste("Proportion for Fatal Collision", prop5))
```
The print statements above detail the distribution of the proportions for each class in the Injury Severity column. There is a clear imbalance that will be corrected via SMOTE.

Tidyverse doesConvert to factor
```{r}
initial_df$injury_severity <- as.factor(initial_df$injury_severity)
```

Will name this data frame balanced_df.
```{r}
# A more balanced dataset
balanced_df <- SMOTE(
  initial_df,
  initial_df$injury_severity
  )

# View distribution of response variable in new dataset
table(balanced_df$injury_severity)
```

The initial dataframe has some simple preprocessing needed to fit a model.
Next, will perform any remaining preprocessing using recipes.
```{r}
library(tidyverse)
#library(ranger)
set.seed(120)  # Setting seed
```
Quick view of the initial data frame.
```{r}
tail(initial_df)
```
A "tail" check reveals that the configuration of the data frame does include the wanted
columns.

Step 1 was transformation of the raw data frame to the initial data frame.
```{r}
df <- raw_data %>%
  filter(
    raw_data$"Weather Condition" != "",
    raw_data$"Lighting Condition" != ""
  ) %>%
  select(
    "City",
    "Weather Condition",
    "Lighting Condition",
    "Injury Severity"
  ) %>%
  # Converting all the character data types to factors
  mutate_if(is.character, as.factor) %>%
  # Change the column names to snake and lower case
  clean_names()
```
So in the above command I incorporated the conversion of character data types to factors
for all the string types.

Simple training/test set split.
```{r}
library(tidymodels)
crash_split <- initial_split(df, prop = 0.8)
train_data <- training(crash_split)
test_data <- testing(crash_split)
```
Get the dimensions of the dataframes.
```{r}
train_data %>% dim()
test_data %>% dim()
```
Want to apply 10-fold cross-validation.
```{r}
train_cv <- vfold_cv(train_data, v = 10)
```
Probably won't use the 10-fold cross-validation.






